A thermodynamic view of dusty protoplanetary disks

### 2017-09-23 paper.tex
tiny edits before second submission

\section{Growing oscillations by doing work}\label{grow_osc}
{\bf
In this section we provide physical arguments to highlight the
similarity between single phase fluids and dust-gas mixtures. It is
well-known that in the limit of perfect coupling the addition of dust
increases the gas inertia but not pressure. One can then regard the
dusty gas as a single mixture with a reduced sound-speed. Here, we
argue that when dust-gas coupling is imperfect, there exists another
similarity with pure gas related the phase of pressure and density
evolution. Thus it is still possible to regard
partially-coupled dusty-gas as a single fluid. The mathematical
description of this section is given in \S\ref{dust_work}.   
}
%dust_work

>>>>

\section{Growing oscillations by doing work}\label{grow_osc}
{\bf
In this section we provide physical arguments to highlight the
similarity between single phase fluids and dust-gas mixtures. It is
well-known that in the limit of perfect coupling the addition of dust
increases the gas inertia but not pressure. One can then regard 
dusty gas as a single fluid with a reduced sound-speed. Here, we
argue that when dust-gas coupling is imperfect, there exists another
similarity with pure gas; related the phase of pressure and density
evolution. Thus it is still possible to regard
partially-coupled dusty-gas as a single fluid. The mathematical
description of this section is given in \S\ref{dust_work}.   
}
%dust_work

### 2017-09-23 summary.tex
tiny edits before second submission

The main application we envision for the dusty/adiabatic gas analogy 
to develop physical interpretations of dust-gas instabilities; and to find 
dusty analogs of pure gas instabilities in protoplanetary disks, 
which is discussed in \S\ref{dust_analogs}. One can also exploit the similarity to adapt existing
hydrodynamic codes to simulate dusty protoplanetary disks (see \S\ref{dust_sims}).

@@

However, the validity of the simplified one-fluid equations may be
problem-dependent, and not just on the terminal velocity 
approximation. For example, Table \ref{si_compare} shows the dust-rich
`linA' mode and dust-poor `linB' mode of the streaming instability
have the same $\tstop$, but the former is accurately captured by the
simplified equations, while the latter is not.  This suggests that for
the streaming instability the simplified equations is better suited
for $\rhod/\rhog>1$. The simplified equations may also contain spurious
modes in certain situations (see \S\ref{spurious_epi}).

@@

\section{Summary}\label{summary}
In this paper we develop the analogy between isothermal 
dusty-gas and pure ideal gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, with stopping time $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy conservation 
following an adiabatic, pure gas.

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also show that radial variations in
  $\rhod/\rhog$ can trigger a type of VSI, even when the 
  usual sources of vertical shear -- vertical dust gradients and radial 
  temperature gradients -- are completely absent.  
  %effectively translates to a non-uniform radial temperature
  %profile, and the resulting vertical shear can lead to instability,
  %even if the true sound-speed is constant. 
}

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_\mathrm{g}$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_g$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop, ][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}

>>>>

The main application we envision for the dusty/adiabatic gas analogy 
is to develop physical interpretations of dust-gas instabilities; and to find 
dusty analogs of pure gas instabilities in protoplanetary disks, 
which is discussed in \S\ref{dust_analogs}. One can also exploit the similarity to adapt existing
hydrodynamic codes to simulate dusty protoplanetary disks (see \S\ref{dust_sims}).

@@

%However, the validity of the simplified one-fluid equations may be
%problem-dependent, and not just on the terminal velocity 
%approximation.

However, the validity of the terminal velocity approximation depends
not just on $\tstop$, but also on the value of the dust gas-ratio, and
possibly the problem itself. 
%You can decide if this is a relevant improvement, and whether to
%include. 
For example, Table \ref{si_compare} shows the dust-rich
`linA' mode and dust-poor `linB' mode of the streaming instability
have the same $\tstop$, but the former is accurately captured by the
simplified equations, while the latter is not.  This suggests that for
the streaming instability the simplified equations is better suited
for $\rhod/\rhog>1$. The simplified equations may also contain spurious
modes in certain limits (see \S\ref{spurious_epi}).

@@

\section{Summary}\label{summary}
{\bf
In this paper, we examine the conditions under which the presence of
dust can trigger instabilities in gaseous protoplanetary disks.  To this end, 
we develop an analogy between isothermal  dusty-gas and pure ideal
gas. 
} 
The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, with stopping time $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy conservation 
following an adiabatic, pure gas.

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also show that radial variations in
  $\rhod/\rhog$ can trigger a type of VSI, even when the 
  usual sources of vertical shear -- vertical dust gradients and radial 
  temperature gradients -- are negligible.  
  %effectively translates to a non-uniform radial temperature
  %profile, and the resulting vertical shear can lead to instability,
  %even if the true sound-speed is constant. 
}

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_\mathrm{g}$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_\mathrm{g}$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop, ][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}
